The reference model approach in feature selection problems \vspace{.1in}

### 2019-06-05 paper/ref_approach.tex
minor fix

In statistical applications, one of the main steps of the modelling workflow is covariate or feature selection, which can be seen as a special case of model reduction. The necessity of carrying out the selection arises from different problematics. Sometimes the main objective of the analysis is doing prediction for future observations, thus the main focus in building the model is on the predictive performance. Yet a high dimensional parameter space often leads to dangers as overfitting or other technical problems, as expensive costs to collect all the features or computational and memory burdens, e.g. when we are operating on limited resources, and thus a sparser representation of the model would be preferred. In addition to that, simpler models bring also advantages in terms of interpretability. Some other times, variable selection can be also seen as an inferential tool: the interest is on knowing which variables are meaningful, or relevant, in relation to the target variable of interest. This typically happens in a \textit{small n large p} scenario, as in many microarray data analysis in medicine or biology. A very large number of genes is usually analysed in two groups of individuals labeled as \textit{positive} or \textit{negative} to some disease. The main goal of the analysis is to spot the subset of explanatory genes in order to perform further experiments \cite[see examples in][]{book:efron}.

@@

In this paper we bring motivations for the use of a reference model. A properly designed model is able to clean part of the noise present in the data, thus to provide an improved and more stable selection process. As an example, we show the improvements in the selection when using the projection predictive approach versus the stepwise regression with a real world dataset. We argue that however the reference model is used, it can be always seen acting as a filter on the observed data. Following this idea, we devise a simple reference model approach that can be applied on top of any selection procedure. We thus show the reference model benefits, regardless of what specific procedure is applied, using as a benchmark the normal means estimation problem. We apply different state-of-the-art methods comparing in each case the results basing the selection on the data or on the reference model. In case of full Bayesian methods, we report results in terms of shrinkage and accuracy of the posterior estimates, whereas when a full selection is carried out, we compute quantities as false discovery rate, sensitivity and a measure of the stability of the selection. The results indicate how the core reason why the reference model based methods perform well is the reference model itself, rather than the specific way of using it. We do recognise that all these methods diversify how the subset of variables is actually chosen and which properties it has, thus it remains fundamental to investigate and develop such different procedures. We hope that this work can increase the attractiveness on this family of methods and, therefore, it can be of interest to a large number of practitioners.
\\
We would like also the reader to note that besides the fact that we consider a Bayesian framework in our discussion, the reference model can and is built also in frequentist settings, as it is done in some of the references previously provided.

@@

Relying on the idea of the reference model as a noise-filter, a simple, but effective, reference model approach which can be used on top of any feature selection procedure consists in substituting the target observations with a point prediction of the reference model, e.g. posterior predictive mean. Afterwards any method for feature selection can be used. In Section \ref{comparison} we compare some well-known methods from the literature using both such reference model approach and the original data, observing at the end of the selection higher stability and quality of the selection thanks to the reference model.

@@

\begin{itemize}
\item add more quantitative results: e.g. times for projpred and steplm
\end{itemize}

We can consider the reference model approach as a family of methods, thus providing an answer to the question what the computational burden is, is not possible, because not unique. It is possible to identify a common and inevitable overhead though, that is the cost of fitting the reference model. In the simple reference model approach we have described at the end of Section \ref{reference-model-approach}, it is actually the only added cost. In general, a full Bayesian model is advisable in order to have the best predictive performance, however including all the available features can be often quite computationally demanding. In such a case a first possible speed-up could be using screening or dimensionality reduction techniques, as for example the supervised principal components \citep{paper:original_spc,paper:piironen_spc}. However, in some cases this does not always result in an appreciable speed-up, as when there are many predictive and not correlated features and thus many of them have to be included in the reference model anyway. Other alternatives could be fixing some parameters with their maximum marginal likelihood value (typically referred as empirical Bayes) or even use frequentist approach, if predictions are good enough.

@@

First we use simulated data in order to vary the difficulty of the selection problem. We rely on the scheme \eqref{eq:simulated_data} using different levels of correlation ($\rho$) and number observation ($n$), precisely $p$ and $k$ are fixed respectively at $1000$ and $100$, whereas $n\in\{50,70,100\}$ and $\rho\in\{0.3,0.5\}$. Lower the correlation level and the number of observation are, more challenging the selection is. As mentioned, this data generation mechanism was already proposed by \cite{paper:projpred}, who also devise a reference model sufficiently good in predictions. It consists in a linear regression using the first five supervised principal components (SPCs) \citep{paper:original_spc, paper:piironen_spc} as follows:
\
\begin{equation}
\label{eq:ref_mod}
\begin{aligned}
    Y_{i}|&\boldsymbol{\beta},\sigma^{2},\boldsymbol{u}_{i} \overset{ind}{\sim} N(\boldsymbol{u}_{i}^{T}\boldsymbol{\beta},\sigma^{2}) \quad &i=1,..,n \\
    &\beta_{j}|\!\begin{aligned}[t] &\tau \overset{iid}{\sim} N(0,\tau^{2})\\
    &\tau \sim t_{4}^{+}(0,s_{max}^{-2}) 
    \end{aligned} &j=1,..,5 \\ 
    &\sigma \sim t_{3}^{+}(0,10) \\
\end{aligned}
\end{equation}
where $\boldsymbol{u}_{i}$ stays for the $i$-th SPC and $s_{max}$ denotes the sample standard deviation of the largest SPC. All model are implemented in Stan \citep{paper:stan}. The normal means problem formulation is obtained as explained in Section \ref{comparison} by Fisher-transformation. We label as ``ref'' the reference approach, which consists in using the reference model on top of the computation of the z-values, and ``data'' the standard approach based on the observed data only. In Section \ref{shrinkage-signal} we analyse in the context of full-Bayes methods the different shrinkage and signal estimation provided by the use of a reference model or not. Section \ref{complete-selection} shows results for the complete selection procedures.

>>>>

In statistical applications, one of the main steps of the modelling workflow is covariate or feature selection, which can be seen as a special case of model reduction. The necessity of carrying out the selection arises from different problematics. Sometimes the leading objective of the analysis is to predict future observations, thus the main focus in building the model is on the predictive performance. Yet a high dimensional parameter space often leads to dangers as overfitting or other technical problems, as expensive costs to collect all the features or computational and memory burdens (e.g. when we are operating on limited resources), and thus a sparser representation of the model would be preferred. In addition to that, simpler models bring also advantages in terms of interpretability. Some other times, variable selection can be also seen as an inferential tool: the interest is on knowing which variables are meaningful, or relevant, in relation to the target variable of interest. This typically happens in a \textit{small n large p} scenario, as in many microarray data analysis in medicine or biology. A very large number of genes is usually analysed in two groups of individuals labeled as \textit{positive} or \textit{negative} to some disease; the goal is to spot the subset of explanatory genes in order to perform further experiments \cite[see examples in][]{book:efron}.

@@

In this paper we bring motivations for the use of a reference model. A properly designed model is able to clean part of the noise present in the data, thus to provide an improved and more stable selection process. As an example, we show the improvements in the selection when using the projection predictive approach versus the stepwise regression, with a real world dataset. We argue that, however the reference model is used, it can be always seen acting as a filter on the observed data. Following this idea, we devise a simple reference model approach that can be applied on top of any selection procedure. We thus show the reference model benefits, regardless of what specific procedure is applied, using as a benchmark the normal means estimation problem. We apply different state-of-the-art methods comparing in each case the results basing the selection on the data or on the reference model. In case of full Bayesian methods, we report results in terms of shrinkage and accuracy of the posterior estimates, whereas when a full selection is carried out, we compute quantities as false discovery rate, sensitivity and a measure of the stability of the selection. The results indicate how the core reason why the reference model based methods perform well is the reference model itself, rather than the specific way of using it. We do recognise that all these methods diversify how the subset of variables is actually chosen and which properties it has, thus it remains fundamental to investigate and develop such different procedures. We hope that this work can increase the attractiveness on this family of methods and, therefore, it can be of interest to a large number of practitioners.
\\
We would like also the reader to note that besides the fact that we consider a Bayesian framework in our discussion, the reference model can and is built also in frequentist settings, as it is done in some of the references previously provided.

@@

Relying on the idea of the reference model as a noise-filter, a simple, but effective, reference model approach which can be used on top of any feature selection procedure consists in substituting the target observations with a point prediction of the reference model, e.g. posterior predictive mean. Afterwards, any method for feature selection can be used. In Section \ref{comparison} we compare some well-known methods from the literature using both such reference model approach and the original data, observing at the end of the selection higher stability and quality of the selection thanks to the reference model.

@@

We can consider the reference model approach as a family of methods, thus providing an answer to the question what the computational burden is, is not possible, because not unique. It is possible to identify a common and inevitable overhead though, that is the cost of fitting the reference model. In the simple reference model approach we have described at the end of Section \ref{reference-model-approach}, it is actually the only added cost. In general, a full Bayesian model is advisable in order to have the best predictive performance, however including all the available features can be often quite computationally demanding. In such a case a first possible speed-up could be using screening or dimensionality reduction techniques, as for example the supervised principal components \citep{paper:original_spc,paper:piironen_spc}. However, in some cases this does not always result in an appreciable speed-up, as when there are many predictive and not correlated features and, thus, many of them have to be included in the reference model anyway. Other alternatives could be fixing some parameters with their maximum marginal likelihood value (typically referred as empirical Bayes) or even use a frequentist approach, if predictions are good enough.

@@

Here we use simulated data in order to control the difficulty of the selection problem. We rely on the scheme \eqref{eq:simulated_data} using different levels of correlation ($\rho$) and number observation ($n$), precisely $p$ and $k$ are fixed respectively at $1000$ and $100$, whereas $n\in\{50,70,100\}$ and $\rho\in\{0.3,0.5\}$. Lower the correlation level and the number of observation are, more challenging the selection is. As mentioned, this data generation mechanism was already proposed by \cite{paper:projpred}, who also devise a reference model sufficiently good in predictions. It consists in a linear regression using the first five supervised principal components (SPCs) \citep{paper:original_spc, paper:piironen_spc} as follows:
\
\begin{equation}
\label{eq:ref_mod}
\begin{aligned}
    Y_{i}|&\boldsymbol{\beta},\sigma^{2},\boldsymbol{u}_{i} \overset{ind}{\sim} N(\boldsymbol{u}_{i}^{T}\boldsymbol{\beta},\sigma^{2}) \quad &i=1,..,n \\
    &\beta_{j}|\!\begin{aligned}[t] &\tau \overset{iid}{\sim} N(0,\tau^{2})\\
    &\tau \sim t_{4}^{+}(0,s_{max}^{-2}) 
    \end{aligned} &j=1,..,5 \\ 
    &\sigma \sim t_{3}^{+}(0,10) \\
\end{aligned}
\end{equation}
where $u_{ij}$ stays for the $j$-th SPC of observation $i$, and $s_{max}$ denotes the sample standard deviation of the largest SPC. All model are implemented in Stan \citep{paper:stan}. The normal means problem formulation is obtained as explained in Section \ref{comparison} by Fisher-transformation. We label as ``ref'' the reference approach, which consists in using the reference model on top of the computation of the z-values, and ``data'' the standard approach based on the observed data only. In Section \ref{shrinkage-signal} we analyse in the context of full-Bayes methods the different shrinkage and signal estimation provided by the use of a reference model or not. Section \ref{complete-selection} shows results for the complete selection procedures.
